Microlensing as a possible probe of event-horizon structure in quasars

### 2017-06-28 paper1/ms.tex
edits

In Section~\ref{sec:source-models} we introduce the three source
profiles used in our subsequent models and simulations: a
constant-brightness disc, a circular Gaussian, and the crescent source
introduced by \cite{2013MNRAS.434..765K}.  The latter is simply a
constant-brightness disc with a smaller, non-concentric disc cut out
of it.  We also derive the half-light radius for a crescent.  The
half-light radius can characterise the source size for all three types
of source.

@@

In Section~\ref{sec:numerics} we carry out source fitting to
lightcurves, with both noise and systematic errors are present.  We
generate a mock lightcurves by taking a crescent source across the
path AB in figure~\ref{fig:magnification_map}, and then adding noise.
The path simulates crossing a clean but not ideal fold.  We then fit
this lightcurve to templates from both crescent and circular-Gaussian
sources across the path CB.  That is, the templates come from a
similar but not identical caustics, thus deliberately generating a
systematic error.  Despite this systematic error, a crescent source
fits the data, while a Gaussian source is rejected from the $\chi^2$
value.

Finally in Section~\ref{sec:discussion} we discuss in more depth the
implications of the results presented in the previously mentioned
sections.  One of the most interesting implication is the possibility
to estimate the black hole mass from the reconstructed
parameters. This further requires good approximations of the relative
transversal velocity between the quasar and the lens. Proper
constraints can be set with independent observations of the stellar
structure that contain the gravitational lens.

@@

Caustics are important in all forms of lensing with multiple images,
but they have a special significance for lens quasars, first pointed
out by \cite{1979Natur.282..561C}.  The granularity of the mass
distribution $\Sigma(\vec\theta)$ due to individual stars produces a
caustic network on the scale of $\sqrt{GM_\odot D_L/c^2}$ or $10^{-6}
\sqrt{M/M_\odot}\rm\,arcsec$ for typical lens and source redshift
\citep{2001PASA...18..207W}.  Extended sources wash out this
micro-caustic structure, but the optical-continuum source of quasars
is even smaller.

@@

For a microlensing event the lightcurve can be written for an
undefined source shape as:
\begin{equation}
 F(t) = \int_{-\infty}^\infty \int_{-\infty}^\infty S_{2D}(p-p_s(t), q-q_s(t)) \mu_t(p) \mathrm{d}q \mathrm{d}p
 \label{eqn:ft2d}
\end{equation}
In order to build the previous equation we have considered that the
time dependency of the flux $F$ is given only by the motion of the
source with respect to a fixed caustic. Therefore the only time
dependent quantities in the right hand side of the equation are the
coordinates of the center of the source $p_s,q_s$ and by construct
$S_{2D}$.

Due to the choice of the coordinate system the amplification factor 
has no dependence on the $q$ coordinate. The previous equation can be rewritten as:

@@

In the present study we are analysing three types of sources with different surface brightness: 
\begin{enumerate}
 \renewcommand{\theenumi}{(\arabic{enumi})}
  \item a rotationally symmetric source with a bivariate gaussian surface brightness distribution,
  \item a disk source with constant surface brightness distribution,
  \item a crescent shaped source with constant surface brightness distribution.
\end{enumerate}
The first two sources are the typical choices used in the literature to describe the luminous parts of a quasar. 
The third one is a recently proposed variant \citep{2013MNRAS.434..765K}.

@@

Although such a definition for a source would have non-zero surface/linear brightness for any coordinate $p,q$, the amount of light received by a detector from outside a $3 \sigma$ disk centered at $p_s, q_s$ 
would be insignificant. For a gaussian distributed surface brightness source the half-light radius is directly proportional to the parameter $\sigma$ according to the equation:
\begin{equation}
r_{1/2} = \sqrt{ln(4)} \sigma.
\end{equation}

@@

One can construct mathematically a disk source with constant surface brightness and radius $R$ using a stepfunction:
\begin{equation}
 S_{2D}^D(p-p_s, q-q_s) = \frac{S_0^D}{\pi R^2} \Theta \left( R^2 - \left( p-p_s \right)^2 - \left( q-q_s \right)^2 \right).
\end{equation}
By integrating over $q$ coordinate the linear brightness function is obtained:

\begin{equation}
 S_{1D}^D(p-p_s) = \frac{2 S_0^D}{\pi R}  \sqrt{1 - \frac{(p-p_s)^2}{R^2} }    \Theta \left( R^2 - \left( p-p_s \right)^2 \right).
\end{equation}
\\
The half-light radius of a uniform disk source is $R/\sqrt{2}$.

@@

\begin{equation}
 S_{2D}^{Dp}(p-p_{sn}, q-q_{sn}) = \frac{S_0^{Dp}}{\pi R_p^2} \Theta \left( R_p^2 - \left( p-p_{sp} \right)^2 - \left( q-q_{sp} \right)^2 \right)
\end{equation}
\\
and
\begin{equation}
 S_{2D}^{Dn}(p-p_{sn}, q-q_{sn}) = \frac{S_0^{Dn}}{\pi R_n^2} \Theta \left( R_n^2 - \left( p-p_{sn} \right)^2 - \left( q-q_{sn} \right)^2 \right).
\end{equation}
\\
The following notations were used: $R_p, (p_{sp}, q_{sp}), R_n, (p_{sn},q_{sn})$ are the radii 
and coordinate of the center for the larger positive disk and smaller negative disk respectively. 
$S_0^{Dp},S_0^{Dn}$ represent the total flux of radiation received from the large and small disk. 
From this point forward we will not use the total flux from each source. Instead we will 
use the difference which in this case is the total flux from the crescent-shaped source $S_0^C$. \\
Equation \ref{eqn:s2d} can be written as:\\

\begin{align}
 S_{2D}^C &= \frac{S_0^C}{\pi \left(R_p^2-R_n^2 \right)} \left\{ \Theta \left[ R_p^2 - \left( p-p_{sp} \right)^2 - \left( q-q_{sp} \right)^2 \right] \right.\nonumber\\
 &\qquad \left. {} -  \Theta \left[ R_n^2 - \left( p-p_{sn} \right)^2 - \left( q-q_{sn} \right)^2 \right] \right\}.
\end{align}
\\
Analogous for the linear brightness function:

@@

There are some constraints on the parameters used to define a crescent in the previously 
presented manner that need to be stated. First, we must impose the obvious $R_p > R_n$ relation. Secondly, 
the small disk must always be interior to the large disk:
\begin{equation}
 R_p \ge R_n + \sqrt{\left(p_{sp} - p_{sn} \right)^2 + \left(q_{sp} - q_{sn} \right)^2}
\end{equation}

For the distances between the centers of the two disks we will use the same 
notations as the one found in the paper \citep{2013MNRAS.434..765K}, $a \equiv p_{sn} - p_{sp}$ and $b \equiv q_{sn} - q_{sp}$.

The half-light radius of any source is invariant to any rotational transformation. 
In the present case of a crescent source the effective radius is dependent on the 
parameters $R_p$, $R_n$ and $\sqrt{a^2+b^2}\equiv c $ exclusively. From symmetry
considerations the centroid of the source is collinear with the centers of the two disks 
and it is situated at a distance $d_c$ from the center of the bright disk. $d_c$ can be 
computed numerically with the use of a variation of equation \ref{eqn:s1_d}:

@@

With the position of the centroid determined, the half-light radius can be also be computed numerically:

@@

Using equation \ref{eqn:ft} and the one-dimensional flux function
presented in the previous section one can compute numerically the
lightcurves of the three extended sources for the simplified
infinite-wall-caustics model.

@@

The amount of light received by an observer from a source 
with a gaussian distributed brightness with $\sigma$ and total flux $S_0^G$ in the absence of any gravitational lensing is:
\begin{equation}
 F^G(t) = \int_{-\infty}^\infty  \left( \mu_0 + \frac{C_0}{\sqrt{p}} \Theta \left( p \right) \right) \left( \frac{S_0^G}{\sqrt{2 \pi} \sigma} e^{-\frac{(p-p_s(t))^2}{2 \sigma^2}} \right) \mathrm{d}p.
\end{equation}

which can be simplified to:
\begin{equation}
 F^G(t) = \mu_0 S_0^G + \frac{C_0 S_0^G}{\sqrt{2\pi} \sigma} \int_{0}^\infty \frac{e^{-\frac{(p-p_s(t))^2}{2 \sigma^2}}}{\sqrt{p}} \mathrm{d}p.
\end{equation}

@@

\begin{equation}
\begin{aligned}
 F^D(t) &= \int_{-\infty}^\infty  \left( \mu_0 + \frac{C_0}{\sqrt{p}} \Theta \left( p \right) \right) \\
    & \bigg[ \frac{2 S_0^D}{ \pi R} \sqrt{1 - \frac{\left( p-p_s(t) \right)^2}{R^2}}  \Theta \left(R^2 - \left(p-p_s(t) \right)^2 \right) \bigg] \mathrm{d}p.
\end{aligned}
\end{equation}

which is equivalent to:
\begin{equation}
 F^D(t) = \mu_0 S_0^D + \frac{2 C_0 S_0^D}{\pi R} \int_{max(0, p_s(t) - R)}^{max(0, p_s(t) + R)} \frac{1}{\sqrt{p}} \sqrt{1 - \frac{\left( p-p_s(t) \right)^2}{R^2}} \mathrm{d}p.
\end{equation}

@@

The function $p_s(t)$ can be chosen to be equal to $v_p(t-t_0) + p_{s0}$. Where $p_{s0}$ is the coordinate $p$ 
of the source at the initial time, and $v_p$ is the component of the velocity
along the $p$ axis. Such a modelling of the motion of the object in the source 
plane describes a linear motion with constant velocity. Furthermore, we reduce the complexity of the model by 
choosing the function $a$ to be constant in time.

@@

In figure \ref{fig:char_points} there are four characteristic points visible on the resulting one-dimensional 
light profile of the crescent shaped source. Two of the points, p1 and p4, mark the outer boundaries of the
 luminous disc component. The other two, p2 and p3, mark the boundaries of the dark disc component. 
Since $S_{1D}$ is a projection of $S_{2D}$ on a line perpendicular to the caustic, the following relation holds:
\begin{equation}
    p_2-p_1 = R_p -R_n - a.
\end{equation}
In addition there are two other obvious relations the characteristic points which are independent of the projection:
\begin{equation}
    p_4 -p_1 = 2 R_p,
\end{equation}

\begin{equation}
        p_3 -p_2 = 2 R_n.
\end{equation}
All four points mark the positions where derivative $\frac{dS_{1D}}{dp}$ is discontinuous. The points can be used 
to define three regions: $p_1 - p_2$ where $S_{1D}$ is convex, $p_2 - p_3$ where $S_{1D}$ is concave, 
and $p_3 - p_4$ where $S_{1D}$ is convex again. \\

@@

The impact of parameter $a$ on the shape of the lightcurve can be
observed in Figure~\ref{fig:a_var}. Sources where the center of the
dark disc reaches the caustic before the center of the bright disc are
characterized by a smoother broad peak in contrast to the cases where
the center of the dark disc reaches the caustic after the bright disk.
In the case of the latter the instance when the dark disc reaches the
caustic corresponds to larger and larger magnifications until it
becomes a local and even global peak. The effect of the $R_n$
parameter on the light curve is presented in
Figures~\ref{fig:lightcurve_crescent} and
\ref{fig:lightcurve_crescent_back}. For the particular set of
parameters where $R_p = R_n +a$ the third period of time discussed
previously does not exist. Particularizing further, if the value of
the radius of the dark disc is comparable to the value of the radius
of the bright one then the position and shape of the maximum
magnification are strikingly different. In case the crescent reaches
the caustic with the bright region first, the peak magnification
happens when the dark region reaches the caustic and it is
characterized by a sharp variation in magnification growth. In the
opposite case, the peak appears before the end of the bright disc
reaches the caustic and the shape of the lightcurve is smoother.

@@

\cite{2012MNRAS.421.1517D} have created a radiative image of M87 based on the GRMHD simulations presented in \citep{2009MNRAS.394L.126M}. 
The top right-most image in figure 5 of \cite{2012MNRAS.421.1517D} has been projected to a 1D profile associated to the perpendicular 
direction to a fold caustic approaching the image from the right. The projection is presented in the upper panel of figure \ref{fig:M87_plots}. 
The amplification values of the flux of light corresponding to a 
microlensing event are presented in the lower panel of figure \ref{fig:M87_plots}. 
In general, the behaviour of the lightcurve is similar to the geometric crescent source with the caveat that the outer regions 
surrounding the luminous parts of the image have non-zero flux and thus are more extended than the simplified source model.

@@

Having used the simple model of an ideal fold to gain insight, we now
consider the question of whether the characteristics of a crescent
source could be discerned in real data having noise and systematic
error as well.  Our method is to generate microlensing lightcurves
using a widely-used microlensing code
\citep{1990PhDT.......180W,1999A&A...346L...5W,1999JCoAM.109..353W} to
which we have added a small extension to provide the option of
crescent sources, and then fit these mock lightcurves to source models
using Markov-chain Monte Carlo (MCMC).  The general strategy is the same
as in previous studies \citep[e.g.,][]{2010ApJ...712..658P} though
the details are different.

@@

The time unit for the light curve is taken to be $r_{1/2}/v_{\!\perp}$
where $v_{\!\perp}$ is the velocity transverse to the caustic.  If we
assume the brightness measurements are nightly, and
$v_{\!\perp}=\rm200\,km/s$, the length of the microlensing event (250
days) is comparable to typical observed events \cite[see
  e.g.,][]{2015ApJ...814L..26M}. The implied $r_{1/2}$ is
$\sim10^{14}\rm\,cm$, which is at the lower end of values inferred for
microlensed X-ray sources, whereas for optical sources, the inferred
$r_{1/2}$ is one to two order of magnitude larger \citep[see, e.g.,
  Figure~8 of][]{0004-637X-769-1-53}.

The mock data appear in Figures
\ref{fig:cc_forward}--\ref{fig:gc_forward} below.  The example values
abovr for $r_{1/2}$, $v_{\!\perp}$ and the observing frequency are not
important.  What is important is that the observations have enough
sampling and signal-to-noise that the characteristic features of a
crescent source (see e.g., Figure~\ref{fig:char_points}) are not
washed out.

@@

We now proceed to fit the mock lightcurve.  The basic fitting
technique is the same as used in   Markov-Chain Monte-Carlo is
used to fit source parameters and additional nuisance parameters (see
below) jointly, and then marginalise out all uninteresting parameters,
leaving posterior distributions for any chosen parameter.

In addition to the source parameters, the fitting procedure includes
three nuisance parameters, namely the brightness normalisation and the
start and end points on the path.  The latter two are equivalent to
considering the location of the caustic with respect to the starting
point and the velocity component of the source transverse to the
caustic (that is, $v_{\!\perp}$).  The orientation of the source track
is another nuisance parameter, which we do not include it, because for
a pure fold, changing the orientation of the source track is
equivalent to changing $v_{\!\perp}$ and the orientation of the source
(see Figure~\ref{fig:infinite_fold}), all parameters that are
included.  Since the caustic is not a perfect fold, disregarding the
orientation of the source track can be expected to introduce some
systematic error.

@@

Figure \ref{fig:cc_forward} shows a test of the fitting procedure.
The computer compares the mock data to different templates
lightcurves.  All the templates come from taking a crescent source
along the path CB in Figure~\ref{fig:magnification_map}, but with
varied crescent parameters and varied start and finish points.  As the
figure shows, the lightcurve can be fitted to within the noise, and
the parameters $r_{1/2}$ and $R_n/R_p$ can be recovered.

>>>>

In Section~\ref{sec:source-models} we introduce three source profiles
which are used in our subsequent models and simulations: a
constant-brightness disc, a circular Gaussian, and the crescent source
introduced by \cite{2013MNRAS.434..765K}.  The latter is simply a
constant-brightness disc with a smaller, non-concentric disc cut out
of it.  We also derive the half-light radius for a crescent.  The
half-light radius can characterise the source size for all three types
of source.

@@

In Section~\ref{sec:numerics} we carry out source fitting to
lightcurves, with both noise and systematic errors are present.  We
generate a mock lightcurve by taking a crescent source across the path
CB in Figure~\ref{fig:magnification_map}, and then adding noise.  The
path simulates crossing a clean but not ideal fold.  We then fit this
lightcurve to templates from both crescent and circular-Gaussian
sources along the paths AB as well as CB.  That is, the templates from
a similar but not identical caustic are also used, thus deliberately
generating a systematic error.  Despite this systematic error, a
crescent source fits the data, while a Gaussian source is rejected
from the $\chi^2$ value.

Finally in Section~\ref{sec:discussion} we discuss in more depth the
implications of the results presented in the previously mentioned
sections.  One of the most interesting implications is the possibility
of estimating the black hole mass from the reconstructed
parameters. This further requires good approximations of the relative
transversal velocity between the quasar and the lens. Proper
constraints can be set with independent observations of the stellar
structure that contain the gravitational lens.

@@

Caustics are important in all forms of lensing with multiple images,
but they have a special significance for lens quasars, first pointed
out by \cite{1979Natur.282..561C}.  The granularity of the mass
distribution $\Sigma(\vec\theta)$ due to individual stars produces a
caustic network on the scale of $\sqrt{GM_\odot D_L/c^2}$ or $10^{-6}
\sqrt{M/M_\odot}\rm\,arcsec$ for typical lens and source redshift
\citep{2001PASA...18..207W}.  Extended sources wash out this
micro-caustic structure, but the optical-continuum and X-ray sources
of quasars are even smaller.

@@

For a microlensing event the lightcurve can be written for an
arbitrary source shape as:
\begin{equation}
 F(t) = \int_{-\infty}^\infty \int_{-\infty}^\infty S_{2D}(p-p_s(t), q-q_s(t)) \mu_t(p) \mathrm{d}q \mathrm{d}p.
 \label{eqn:ft2d}
\end{equation}
In order to build the previous equation we have considered that the
time dependency of the flux $F$ is given only by the motion of the
source with respect to a fixed caustic. Therefore the only time
dependent quantities in the right hand side of the equation are the
coordinates of the center of the source $p_s(t),q_s(t)$.

Due to the choice of coordinate system the amplification factor 
has no dependence on the $q$ coordinate. The previous equation can be rewritten as:

@@

In the present study we analyse three types of sources with different surface brightness: 
\begin{enumerate}
 \renewcommand{\theenumi}{(\arabic{enumi})}
  \item a rotationally symmetric source with a bivariate gaussian surface brightness distribution,
  \item a disk source with constant surface brightness distribution,
  \item a crescent shaped source with constant surface brightness distribution.
\end{enumerate}
The first two sources are the typical choices used in the literature to describe the luminous parts of a quasar. 
The third one is a recently proposed variant \citep{2013MNRAS.434..765K}.

@@

Although such a definition for a source would have non-zero surface/linear brightness for any coordinate $p,q$, the amount of light received by a detector from outside a $3 \sigma$ disk centered at $p_s, q_s$ 
would be insignificant. For a gaussian distributed surface brightness source the half-light radius is directly proportional to the parameter $\sigma$ according to the equation:
\begin{equation}
r_{1/2} = \sqrt{\ln4}\,\sigma.
\end{equation}

@@

One can construct mathematically a disk source with constant surface-brightness and radius $R$ using a stepfunction:
\begin{equation}
 S_{2D}^D(p-p_s, q-q_s) = \frac{S_0^D}{\pi R^2} \Theta \left( R^2 - \left( p-p_s \right)^2 - \left( q-q_s \right)^2 \right).
\end{equation}
By integrating over $q$ coordinate the linear brightness function is obtained:

\begin{equation}
 S_{1D}^D(p-p_s) = \frac{2 S_0^D}{\pi R}  \sqrt{1 - \frac{(p-p_s)^2}{R^2} }    \Theta \left( R^2 - \left( p-p_s \right)^2 \right).
\end{equation}
The half-light radius of a uniform disk source is $R/\sqrt{2}$.

@@

\begin{equation}
 S_{2D}^{Dp}(p-p_{sn}, q-q_{sn}) = \frac{S_0^{Dp}}{\pi R_p^2} \Theta \left( R_p^2 - \left( p-p_{sp} \right)^2 - \left( q-q_{sp} \right)^2 \right)
\end{equation}
\\
and
\begin{equation}
 S_{2D}^{Dn}(p-p_{sn}, q-q_{sn}) = \frac{S_0^{Dn}}{\pi R_n^2} \Theta \left( R_n^2 - \left( p-p_{sn} \right)^2 - \left( q-q_{sn} \right)^2 \right).
\end{equation}
The following notations were used: $R_p$ and $(p_{sp}, q_{sp})$ are
the radii and coordinate of the center for the larger positive disk, 
while $R_n$ $(p_{sn},q_{sn})$ correspond to the smaller negative disk. 
$S_0^{Dp},S_0^{Dn}$ represent the total flux of radiation received from the large and small disk. 
From this point forward we will not use the total flux from each source. Instead we will 
use the difference which in this case is the total flux from the crescent-shaped source $S_0^C$. \\
Equation \ref{eqn:s2d} can be written as:\\

\begin{align}
 S_{2D}^C &= \frac{S_0^C}{\pi \left(R_p^2-R_n^2 \right)} \left\{ \Theta \left[ R_p^2 - \left( p-p_{sp} \right)^2 - \left( q-q_{sp} \right)^2 \right] \right.\nonumber\\
 &\qquad \left. {} -  \Theta \left[ R_n^2 - \left( p-p_{sn} \right)^2 - \left( q-q_{sn} \right)^2 \right] \right\}.
\end{align}
\\
Analogously for the linear brightness function:

@@

There are some constraints on the parameters used to define a crescent
in the previously presented manner that need to be stated. First, we
must impose the obvious $R_p \geq R_n$ relation. Secondly, the small disk
must always be interior to the large disk:
\begin{equation}
 R_p \ge R_n + \sqrt{\left(p_{sp} - p_{sn} \right)^2 + \left(q_{sp} - q_{sn} \right)^2}
\end{equation}
For the distances between the centers of the two disks we will use the same 
notations as the one found in the paper \citep{2013MNRAS.434..765K}, $a \equiv p_{sn} - p_{sp}$ and $b \equiv q_{sn} - q_{sp}$.

The half-light radius of any source is invariant under any rotational
transformation.  In the present case of a crescent source the
effective radius is dependent on the parameters $R_p$, $R_n$ and
$\sqrt{a^2+b^2}\equiv c $ exclusively. From symmetry considerations
the centroid of the source is collinear with the centers of the two
disks and it is situated at a distance $d_c$ from the center of the
bright disk. $d_c$ can be computed numerically with the use of a
variation of equation \ref{eqn:s1_d}:

@@

With the position of the centroid determined, the half-light radius
can be also be computed through a numerical integral:

@@

Using equation \ref{eqn:ft} and the one-dimensional flux function
presented in the previous section one can numerically compute the
lightcurves of the three extended sources for the simplified
infinite-wall-caustics model.

@@

The amount of light received by an observer from a source 
with a gaussian distributed brightness with $\sigma$ and total flux $S_0^G$ in the absence of any gravitational lensing is:
\begin{equation}
 F^G(t) = \int_{-\infty}^\infty  \left( \mu_0 + \frac{C_0}{\sqrt{p}} \Theta \left( p \right) \right) \left( \frac{S_0^G}{\sqrt{2 \pi} \sigma} e^{-\frac{(p-p_s(t))^2}{2 \sigma^2}} \right) \mathrm{d}p.
\end{equation}
which can be simplified to:
\begin{equation}
 F^G(t) = \mu_0 S_0^G + \frac{C_0 S_0^G}{\sqrt{2\pi} \sigma} \int_{0}^\infty \frac{e^{-\frac{(p-p_s(t))^2}{2 \sigma^2}}}{\sqrt{p}} \mathrm{d}p.
\end{equation}

@@

\begin{equation}
\begin{aligned}
 F^D(t) &= \int_{-\infty}^\infty  \left( \mu_0 + \frac{C_0}{\sqrt{p}} \Theta \left( p \right) \right) \\
    & \bigg[ \frac{2 S_0^D}{ \pi R} \sqrt{1 - \frac{\left( p-p_s(t) \right)^2}{R^2}}  \Theta \left(R^2 - \left(p-p_s(t) \right)^2 \right) \bigg] \mathrm{d}p.
\end{aligned}
\end{equation}
which is equivalent to:
\begin{equation}
 F^D(t) = \mu_0 S_0^D + \frac{2 C_0 S_0^D}{\pi R} \int_{max(0, p_s(t) - R)}^{max(0, p_s(t) + R)} \frac{1}{\sqrt{p}} \sqrt{1 - \frac{\left( p-p_s(t) \right)^2}{R^2}} \mathrm{d}p.
\end{equation}

@@

The function $p_s(t)$ can be chosen to be equal to $v_p(t-t_0) + p_{s0}$ where $p_{s0}$ is the coordinate $p$ 
of the source at the initial time, and $v_p$ is the component of the velocity
along the $p$ axis. Such a model of the motion of the object in the source 
plane describes a linear motion with constant velocity. Furthermore, we reduce the complexity of the model by 
choosing the function $a$ to be constant in time.

@@

In Figure \ref{fig:char_points} there are four characteristic points visible on the resulting one-dimensional 
light profile of the crescent shaped source. Two of the points, p1 and p4, mark the outer boundaries of the
 luminous disc component. The other two, p2 and p3, mark the boundaries of the dark disc component. 
Since $S_{1D}$ is a projection of $S_{2D}$ on a line perpendicular to the caustic, the following relation holds:
\begin{equation}
    p_2-p_1 = R_p -R_n - a.
\end{equation}
In addition there are two other obvious relations the characteristic points which are independent of the projection:
\begin{equation}
    p_4 -p_1 = 2 R_p,
\end{equation}

\begin{equation}
        p_3 -p_2 = 2 R_n.
\end{equation}
All four points mark the positions where the derivative $\frac{dS_{1D}}{dp}$ is discontinuous. The points can be used 
to define three regions: $p_1 - p_2$ where $S_{1D}$ is convex, $p_2 - p_3$ where $S_{1D}$ is concave, 
and $p_3 - p_4$ where $S_{1D}$ is convex again. \\

@@

The impact of parameter $a$ on the shape of the lightcurve can be
observed in Figure~\ref{fig:a_var}. Sources where the center of the
dark disc reaches the caustic before the center of the bright disc are
characterized by a smoother broad peak in contrast to the cases where
the center of the dark disc reaches the caustic after the bright disk.
In the case of the latter the instance when the dark disc reaches the
caustic corresponds to larger and larger magnifications until it
becomes a local and even global peak. The effect of the $R_n$
parameter on the light curve is presented in
Figures~\ref{fig:lightcurve_crescent_back} and
\ref{fig:lightcurve_crescent}. For the particular set of
parameters where $R_p = R_n +a$ the third period of time discussed
previously does not exist. Particularizing further, if the value of
the radius of the dark disc is comparable to the value of the radius
of the bright one then the position and shape of the maximum
magnification are strikingly different. In case the crescent reaches
the caustic with the bright region first, the peak magnification
happens when the dark region reaches the caustic and it is
characterized by a sharp variation in magnification growth. In the
opposite case, the peak appears before the end of the bright disc
reaches the caustic and the shape of the lightcurve is smoother.

@@

\cite{2012MNRAS.421.1517D} have created a radiative image of M87 based on the GRMHD simulations presented in \citep{2009MNRAS.394L.126M}. 
The top right-most image in figure 5 of \cite{2012MNRAS.421.1517D} has been projected to a 1D profile associated to the perpendicular 
direction to a fold caustic approaching the image from the right. The projection is presented in the upper panel of Figure \ref{fig:M87_plots}. 
The amplification values of the flux of light corresponding to a 
microlensing event are presented in the lower panel of Figure \ref{fig:M87_plots}. 
In general, the behaviour of the lightcurve is similar to the geometric crescent source with the caveat that the outer regions 
surrounding the luminous parts of the image have non-zero flux and thus are more extended than the simplified source model.

@@

Having used the simple model of an ideal fold to gain insight, we now
consider the question of whether the characteristics of a crescent
source could be discerned in real data having noise and systematic
error as well.  Our method is to generate microlensing lightcurves
using a widely-used microlensing code
\citep{1990PhDT.......180W,1999A&A...346L...5W,1999JCoAM.109..353W} to
which we have added a small extension to provide the option of
crescent sources, and then fit these mock lightcurves to source models
using Markov-chain Monte Carlo (MCMC).  The basic technique is the same
as in previous studies \citep[e.g.,][]{2010ApJ...712..658P} though
the details are different.

@@

The time unit for the light curve is taken to be $r_{1/2}/v_p$
where $v_p$ is the velocity transverse to the caustic.  If we
assume the brightness measurements are nightly, and
$v_p=\rm200\,km/s$, the length of the microlensing event (250
days) is comparable to typical observed events \cite[see
  e.g.,][]{2015ApJ...814L..26M}. The implied $r_{1/2}$ is
$\sim10^{14}\rm\,cm$, which is at the lower end of values inferred for
microlensed X-ray sources, whereas for optical sources, the inferred
$r_{1/2}$ is one to two order of magnitude larger \citep[see, e.g.,
  Figure~8 of][]{0004-637X-769-1-53}.

The mock data appear in Figures
\ref{fig:cc_forward}--\ref{fig:gc_forward}.  The example values
above for $r_{1/2}$, $v_p$ and the observing frequency are not
important.  What is important is that the observations have enough
sampling and signal-to-noise that the characteristic features of a
crescent source (see e.g., Figure~\ref{fig:char_points}) are not
washed out.

@@

We now proceed to fit the mock lightcurve.  MCMC is used to fit source
parameters and additional nuisance parameters (see below) jointly, and
then marginalise out all uninteresting parameters, leaving posterior
distributions for any chosen parameter.

In addition to the source parameters, the fitting procedure includes
three nuisance parameters, namely the brightness normalisation and the
start and end points on the path.  The latter two are equivalent to
considering the location of the caustic with respect to the starting
point and the velocity component of the source transverse to the
caustic (that is, $v_p$).  The orientation of the source track is
another nuisance parameter, but we do not include it, because for a
pure fold, changing the orientation of the source track is equivalent
to changing $v_p$ and the orientation of the source (see
Figure~\ref{fig:infinite_fold}), all parameters that are included.
Since the caustic is not a perfect fold, disregarding the orientation
of the source track can be expected to introduce some systematic
error.

@@

Figure \ref{fig:cc_forward} shows a test of the fitting procedure.
The computer compares the mock data to different template
lightcurves. The templates come from taking a crescent source along
the path CB in Figure~\ref{fig:magnification_map}, but with varied
crescent parameters and varied start and finish points.  As the figure
shows, the lightcurve can be fitted to within the noise, and the
parameters $r_{1/2}$ and $R_n/R_p$ can be recovered.
